A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-08 tex/codatadp.tex
removed footnotes from conf version

\def\thewordpaper{paper}

\begin{report}
\def\thewordpaper{report}
\end{report}

@@

% \paragraph{Related Work.}
% 
%     * point to Barrett et al. for SMT datatypes
%       * about their own work, they say: "our focus is on generality and
%         efficiency rather than immediacy of implementation"
%       * (deal more directly with finite sorts than Barrett et al., Section 6.1)
%     * SPASS-Pirate
%     * additional ones since then (e.g. strings?)
%     * anything about codatatypes?
%       * proof assistants like Agda, Coq, etc. have them
%       * Dafny, CoALP
%       * also a lot of theoretical research, some of which is loosely connected,
%         e.g. decision procedure for corecursive functions (Henning in Nijmegen)
% 
%     * ODDITY: Oppen 1980: single-constructor, recursive -- infinite values?

@@

A (co)datatype specification consists of $l$~mutually recursive types which are
either all datatypes or all codatatypes. Polymorphic types, nested
(co)recursion, and datatype--codatatype mixtures fall outside this fragment.%
\footnote{%
In principle, rank-1 (top-level) polymorphism \cite{blanchette-paskevich-2013}
should not raise any special difficulties. Nesting datatypes inside datatypes,
and likewise for codatatypes, can be reduced to the mutual case
\cite{gunter-1993-not}. So the only genuinely interesting cases missing are
mixed nested (co)recursion as well as
(co)recursion through a non-(co)datatype (both of which make sense
\cite{blanchette-et-al-2014-codata}).}
In the presentation, we allow ourselves some metalevel parameterization
% at the metalevel
by using subscripts---for example, $\ty{stream}_{\,\tau}$ denotes a
family of ground types including
$\ty{stream}_{\,\ty{int}}$, $\ty{stream}_{\,\ty{bool}}$,
and \smash{$\ty{stream}_{\,\ty{stream}_{\,\ty{real}}}$}.

Each datatype $\delta$ is equipped with
$m \ge 1$ constructors, and each constructor for $\delta$ takes zero or more
arguments and returns a $\delta$ value. The argument types must be either
ordinary, among the already know (co)datatypes, or among the (co)datatypes being
introduced.
%
To every argument corresponds a selector. The introduced names for the (co)data\-types, the
constructors, and the selectors must be distinct and different from already
introduced ones.%
\footnote{It can be convenient to specify the same selector
for several constructors associated with the same (co)data\-type,
as long as the argument types coincide. % \cite[Section~3]{blanchette-et-al-2014-codata}.
However, this is disallowed by SMT-LIB 2, so we do not consider it here.}
Schematically:
%
\[
\begin{aligned}[t]
(\keyw{co})\keyw{datatype}\;\,
  \delta_1 & {}=\, \const{C}_{1}\rlap{${}_{1}$}\phantom{{}_{m}}(\bigl[\const{s}_{11}^1{:}\bigr]\; \tau_{11}^1, \,\ldots,\, \bigl[\const{s}_{11}^{n_{11}}{:}\bigr]\; \tau_{11}^{n_{11}}) & \\
           & {}\;\phantom{=\,}\;{} \smash{\,\,\,\vdots} \\[-\jot] % {}\;\phantom{=}\llap{\ensuremath{\mid}}\; \cdots \\
           & {}\;\phantom{=\,}\llap{\ensuremath{\,\mid\,}}\; \const{C}_{1m}(\bigl[\const{s}_{1m}^1{:}\bigr]\; \tau_{1m}^1, \,\ldots,\, \big[\const{s}_{1m}^{n_{1m}}{:}\big]\; \tau_{1m}^{n_{1m}}) \\[-1\jot]
   \smash{\vdots\,\,\,} \\[-1\jot]
  \keyw{and}\; \,\delta_l & =\, \ldots
\end{aligned}
\]
%
with
$\const{C}_{i\negvthinspace j} : \tau_{i\negvthinspace j}^1\times\cdots\times\tau_{i\negvthinspace j}^{\,k_{\smash{i\negvthinspace j}}} \to \delta_i$
and $\const{s}_{i\negvthinspace j}^{\,k} : \delta_i \to \tau_{i\negvthinspace j}^{\,k}$. Defaults are assumed for
the selector names if they are omitted.

>>>>

\def\thewordpaper{paper}
\newcommand\dotReportFootnote[1]{.}

\begin{report}
\def\thewordpaper{report}
\newcommand\dotReportFootnote[1]{.\footnote{#1}}
\end{report}

@@

\paragraph{Related Work.}
Although it was written several years ago, the account of related work in
Barrett et al.\ \cite{barrett-et-al-2007} is still largely up to date.
Since then, datatypes have been added to Z3 in unpublished work by Leonardo de
Moura and to a SPASS-like prototype prover called Pirate by Daniel Wand.
%\cite{wand-weidenbach-201x}.
Closely related is the support for structural
induction in CVC4 by Reynolds and Kuncak \cite{reynolds-kuncak-2015} and in
Pirate by Wand and Weidenbach \cite{wand-weidenbach-201x}, which depends on a
notion of datatype.

%     * point to Barrett et al. for SMT datatypes
%       * about their own work, they say: "our focus is on generality and
%         efficiency rather than immediacy of implementation"
%       * (deal more directly with finite sorts than Barrett et al., Section 6.1)
%     * SPASS-Pirate
%     * additional ones since then (e.g. strings?)
%     * anything about codatatypes?
%       * proof assistants like Agda, Coq, etc. have them
%       * Dafny, CoALP
%       * also a lot of theoretical research, some of which is loosely connected,
%         e.g. decision procedure for corecursive functions (Henning in Nijmegen)
% 
%     * ODDITY: Oppen 1980: single-constructor, recursive -- infinite values?

@@

A (co)datatype specification consists of $l$~mutually recursive types which are
either all datatypes or all codatatypes. Polymorphic types, nested
(co)recursion, and datatype--codatatype mixtures fall outside this fragment.%
\dotReportFootnote{%
In principle, rank-1 (top-level) polymorphism \cite{blanchette-paskevich-2013}
should not raise any special difficulties. Nesting datatypes inside datatypes,
and likewise for codatatypes, can be reduced to the mutual case
\cite{gunter-1993-not}. So the only genuinely interesting cases missing are
mixed nested (co)recursion as well as
(co)recursion through a non-(co)datatype (both of which make sense
\cite{blanchette-et-al-2014-codata}).}
In the presentation, we allow ourselves some metalevel parameterization
% at the metalevel
by using subscripts---for example, $\ty{stream}_{\,\tau}$ denotes a
family of ground types including
$\ty{stream}_{\,\ty{int}}$, $\ty{stream}_{\,\ty{bool}}$,
and \smash{$\ty{stream}_{\,\ty{stream}_{\,\ty{real}}}$}.

Each datatype $\delta$ is equipped with
$m \ge 1$ constructors, and each constructor for $\delta$ takes zero or more
arguments and returns a $\delta$ value. The argument types must be either
ordinary, among the already know (co)datatypes, or among the (co)datatypes being
introduced.
%
To every argument corresponds a selector. The introduced names for the (co)data\-types, the
constructors, and the selectors must be distinct and different from already
introduced ones%
\dotReportFootnote{It can be convenient to specify the same selector
for several constructors associated with the same (co)data\-type,
as long as the argument types coincide. % \cite[Section~3]{blanchette-et-al-2014-codata}.
However, this is disallowed by SMT-LIB 2, so we do not consider it here.}
Schematically:
%
\[
\begin{aligned}[t]
(\keyw{co})\keyw{datatype}\;\,
  \delta_1 & {}=\, \const{C}_{1}\rlap{${}_{1}$}\phantom{{}_{m}}(\bigl[\const{s}_{11}^1{:}\bigr]\; \tau_{11}^1, \,\ldots,\, \bigl[\const{s}_{11}^{n_{11}}{:}\bigr]\; \tau_{11}^{n_{11}}) & \\
           & {}\;\phantom{=\,}\;{} \smash{\,\,\,\vdots} \\[-\jot] % {}\;\phantom{=}\llap{\ensuremath{\mid}}\; \cdots \\
           & {}\;\phantom{=\,}\llap{\ensuremath{\,\mid\,}}\; \const{C}_{1m}(\bigl[\const{s}_{1m}^1{:}\bigr]\; \tau_{1m}^1, \,\ldots,\, \big[\const{s}_{1m}^{n_{1m}}{:}\big]\; \tau_{1m}^{n_{1m}}) \\[-1\jot]
   \smash{\vdots\,\,\,} \\[-1\jot]
  \keyw{and}\; \,\delta_l & =\, \ldots
\end{aligned}
\]
%
with
$\const{C}_{i\negvthinspace j} : \tau_{i\negvthinspace j}^1\times\cdots\times\tau_{i\negvthinspace j}^{\,k_{\smash{i\negvthinspace j}}} \to \delta_i$
and $\const{s}_{i\negvthinspace j}^{\,k} : \delta_i \to \tau_{i\negvthinspace j}^{\,k}$. Defaults are assumed for
the selector names if they are omitted.
